---
output:
  github_document:
    toc: false
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 9,
  fig.height = 4,
  out.width = "100%",
  dev = "svglite",
  message = FALSE,
  warning = FALSE
)
```

# cppDE

<!-- badges: start -->
[![R-CMD-check](https://github.com/dModverse/cppDE/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dModverse/cppDE/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**Automated C++ code generation for ODE integration with parameter
sensitivities.**

cppDE writes, compiles and runs a C++ solver for a system of ordinary
differential equations given as symbolic equations. First- and second-order
parameter sensitivities are computed in the same integration by automatic
differentiation.

## Features

- **Solvers**: variable-order BDF/NDF and Adams multistep in Nordsieck form,
  adapted from [SUNDIALS/CVODE](https://computing.llnl.gov/projects/sundials);
  Rosenbrock4 extended from
  [Boost.Odeint](https://www.boost.org/doc/libs/release/libs/numeric/odeint/);
  Tsit5, a 7-stage explicit FSAL Runge-Kutta with dense output
- **Sensitivities in both directions**: forward mode carries one direction per
  parameter through the solve, reverse mode contracts a seeded functional in one
  backward sweep at a cost independent of the parameter count
- **Second order**: exact Hessians, forward over forward or forward over reverse
- **Events**: time-based and root-triggered, with saltation corrections that
  keep first- and second-order sensitivities continuous across a jump
- **Symbolic Jacobian**, derived and compiled as analytic C++, with automatic
  sparsity detection and KLU factorisation for large systems
- **`cppFUN()`**: the same code generation for plain multivariate functions,
  with analytic Jacobian, Hessian and vector-Jacobian product
- **Batched solving**: `solveODEBatch()` integrates many conditions in one
  `.Call` over OpenMP threads, avoiding the fork and serialization a
  `parallel::mclapply()` loop pays per evaluation

## Installation

```r
# install.packages("devtools")
devtools::install_github("dModverse/cppDE")
```

Model code is generated and compiled at run time, so a **C++17 compiler** and
**Python with SymPy** are required. Windows users need
[Rtools](https://cran.r-project.org/bin/windows/Rtools/).

### Optional system dependencies

The package installs and runs without any system library. Three features are
gated on one, detected by `./configure` at install time:

| Feature | Needs | Without it |
|:--|:--|:--|
| `cvode()` backend | SUNDIALS (>= 6.0) | the native solvers are unaffected |
| sparse Jacobian path | SuiteSparse / KLU | `sparse = TRUE` reports how to enable itself |
| parallel `solveODEBatch()` | a toolchain with OpenMP | correct results, conditions run serially |

They are independent: KLU is not part of SUNDIALS, and `cppODE()` calls KLU
without involving SUNDIALS at all. OpenMP needs nothing installed on Linux or
Rtools; on macOS, Apple's clang needs `libomp` (`brew install libomp`).

Either install the libraries as system packages,

| Platform | SUNDIALS | SuiteSparse / KLU |
|:--|:--|:--|
| Debian/Ubuntu | `sudo apt install libsundials-dev` | `sudo apt install libsuitesparse-dev` |
| Fedora | `sudo dnf install sundials-devel` | `sudo dnf install suitesparse-devel` |
| Arch | `sudo pacman -S sundials` | `sudo pacman -S suitesparse` |
| macOS (Homebrew) | `brew install sundials` | `brew install suite-sparse` |

or let cppDE build pinned releases into a per-user cache, which needs no
administrator rights:

```r
cppDE::install_libs("sundials")        # or "suitesparse"
```

`./configure` scans that cache on every run, so the build is needed once and
later re-installs find it by themselves. See `?install_libs` for building
during the install itself, for build options, and for removing the cache again.
`CPPDE_EXTRA_CXXFLAGS` is appended to the flags of every model `compile()`.

To see what was detected:

```r
writeLines(readLines(system.file("cvodeConfig.dcf", package = "cppDE")))
```

## Fitting measured magnetisation dynamics

Fit of a Landau-Lifshitz-Gilbert model to the pump-probe measurements of
Zhang et al. (2025). cppDE integrates the model and its parameter
sensitivities. Likelihood, noise model and optimiser are R functions from
[`inst/examples/fitTools.R`](inst/examples/fitTools.R). The code chunks are the
sections of
[`inst/examples/example_fitCoPt.R`](inst/examples/example_fitCoPt.R).

```{r example-sections, include = FALSE}
## Chunks below are read from the numbered sections of the example script.
example <- readLines("inst/examples/example_fitCoPt.R")
banner  <- grep("^## -{20,}$", example)
first   <- banner[c(TRUE, FALSE)]
last    <- c(first[-1] - 1, length(example))
section <- lapply(seq_along(first), function(i) {
  code <- example[(first[i] + 3):last[i]]
  code[cumsum(nzchar(code)) > 0 & rev(cumsum(rev(nzchar(code)))) > 0]
})
names(section) <- sub("^## ([0-9]+)\\..*", "\\1", example[first + 1])
```

### The experiment

| | |
|:--|:--|
| sample | [Co (0.5 nm) / Pt (0.9 nm)]<sub>5</sub>, perpendicular magnetic anisotropy |
| pump | 80 fs, 1500 nm |
| probe | anomalous Hall conductivity, read with a THz pulse, delays up to 80 ps |
| field | 1 to 3 T, seven values, about 30° from the film normal |

The signal rises within 0.5 ps (ultrafast demagnetisation) and recovers within
a few picoseconds. The recovery carries a precession at 40 to 100 GHz. The
seven traces are fitted jointly.

```{r setup-fit}
library(cppDE)
library(ggplot2)
source(system.file("examples", "fitTools.R", package = "cppDE"))

theme_set(theme_bw(base_size = 11))
cores <- if (.Platform$OS.type == "windows") 1L else parallel::detectCores()
```

```{r data, code = section[["1"]]}
```

### Model

The magnetisation direction $\mathbf m$, $|\mathbf m| = 1$, follows the
Landau-Lifshitz-Gilbert equation in Landau-Lifshitz form,

$$
\dot{\mathbf m} = -\frac{\gamma}{1+\alpha^2}
\left[\mathbf m\times\mathbf h + \alpha\,\mathbf m\times(\mathbf m\times\mathbf h)\right],
\qquad \gamma = \frac{g\,\mu_B}{\hbar},
$$

with Gilbert damping $\alpha$ and g-factor $g$. The effective field is the
applied field, tilted by $\phi$ from the normal $\hat{\mathbf z}$, plus the
perpendicular anisotropy field $H_a$,

$$
\mathbf h = H\,(\sin\phi,\,0,\,\cos\phi) + H_a\,\bigl(1 - k(t)\bigr)\,m_z\,\hat{\mathbf z}.
$$

Heating lowers the anisotropy field by the fraction $k(t) = K\,\kappa(t)$,
$\kappa(0) = 1$, which decays as $\dot\kappa = -\kappa/\tau_K$.

Before the pump, $\mathbf m$ rests at
$\mathbf m_0 = (\sin\theta_0, 0, \cos\theta_0)$, where the torque vanishes,

$$
H\sin(\theta_0-\phi) + H_a\sin\theta_0\cos\theta_0 = 0 ,
$$

and the field there is parallel to $\mathbf m_0$,
$\mathbf h(\mathbf m_0) = \lambda\,\mathbf m_0$ with
$\lambda = H\cos(\theta_0-\phi) + H_a\cos^2\theta_0$. The drop of $H_a$
moves the equilibrium; $\mathbf m$ precesses about it while $H_a$ recovers.

The deviation from $\mathbf m_0$ is of order $K$, and the states of the model
are $\boldsymbol\xi = (\mathbf m - \mathbf m_0)/K$ and $\kappa$. Because
$\mathbf m_0\times\mathbf h(\mathbf m_0) = 0$, the right-hand side in these
states is a polynomial in $K$ without a constant term. The equations hold for
any $K$, and $K\to 0$ gives the linearised dynamics. This avoids computing
$m_{z0} - m_z$ as a difference of nearly equal numbers.

```{r model, code = section[["2"]]}
```

### Observable, sensitivities and noise

The anomalous Hall conductivity is proportional to the out-of-plane
magnetisation. The recovery of the demagnetisation is not part of the model and
is described per measurement by two exponentials:

$$
y(t) = A_1 e^{-t/\tau_1} + A_2 e^{-t/\tau_2} + S\,\bigl[m_{z0} - m_z(t)\bigr]
     = A_1 e^{-t/\tau_1} + A_2 e^{-t/\tau_2} - S\,K\,\xi_z(t).
$$

$\alpha$, $g$, $H_a$, $\tau_K$, $S$, $K$ and $\phi$ are shared by all fields;
background and noise level are per measurement. $H$ is the set field value;
$\mu_B/\hbar$ is the only fixed constant.

cppDE returns $\partial\boldsymbol\xi/\partial p$ for the parameters of the
model. $\theta_0$ and $\lambda$ depend on $H_a$ and $\phi$; the implicit
function theorem gives

$$
\frac{\partial\theta_0}{\partial H_a} = -\frac{\sin\theta_0\cos\theta_0}{D},\qquad
\frac{\partial\theta_0}{\partial\phi} = \frac{H\cos(\theta_0-\phi)}{D},\qquad
D = H\cos(\theta_0-\phi) + H_a\cos 2\theta_0 ,
$$

and the chain rule gives the sensitivities with respect to $H_a$ and $\phi$.
No finite differences are used.

The autocorrelation of the residuals decays within five samples, about one
picosecond. The noise is modelled as a moving average (Box et al. 2015),

$$
u_t = \eta_t + \sum_{j=1}^{5} w_j\,\eta_{t-j},\qquad \eta_t\sim\mathcal N(0,\sigma^2),
$$

with the kernel $w$ shared by all measurements and $\sigma$ per measurement.
Correlations beyond five samples are zero, so the noise model cannot describe
the precession. The likelihood is the conditional sum of squares of the
innovations $\eta_t$; the $\log\sigma$ term enters as a pseudo-residual, which
keeps the problem least squares.

```{r likelihood, code = section[["3"]]}
```

### Fit

Gauss-Newton on a trust region with exact subproblem (More-Sorensen), from
100 random starts. Positive parameters are fitted as $\log_{10}$, the fractions
$K$ and $\phi/90^\circ$ as $\log_{10}\bigl(x/(1-x)\bigr)$, so one unit is one
decade. The physics parameters carry a Gaussian prior with a standard deviation
of three decades, centred on orders of magnitude and the nominal field angle.

```{r multistart, code = section[["4"]]}
```

```{r fit, code = section[["5"]]}
```

The innovations pass the Ljung-Box test (Ljung and Box 1978). For small precession amplitudes $S$,
$K$ and $\phi$ enter essentially as a product and are not determined
individually.

Small-angle precession frequency about the equilibrium (Smit-Beljers) against
the frequencies obtained in the paper from damped-sine fits:

```{r frequencies, code = section[["7"]]}
```

Data: measured traces of Fig. 3 in Z. Zhang, Y. Shiota, S. Karube,
Y. Watanabe, T. Ono and H. Hirori,
[Phys. Rev. Applied **24**, 044079 (2025)](https://doi.org/10.1103/4p39-gyrd),
published under CC BY 4.0 in
[doi:10.5281/zenodo.17333041](https://doi.org/10.5281/zenodo.17333041); see
[`inst/COPYRIGHTS`](inst/COPYRIGHTS).

Noise model: G. E. P. Box, G. M. Jenkins, G. C. Reinsel and G. M. Ljung,
*Time Series Analysis: Forecasting and Control*, 5th ed. (Wiley, 2015).
G. M. Ljung and G. E. P. Box, *On a measure of lack of fit in time series
models*, [Biometrika **65**, 297 (1978)](https://doi.org/10.1093/biomet/65.2.297).

## License

cppDE is distributed under the **MIT License**; see `LICENSE` / `LICENSE.md`
for the full text.

Portions of the C++ solver core in `inst/include/cppde/` are derived from
third-party projects and remain subject to their upstream licenses:

- The Nordsieck multistep stepper (BDF/NDF and Adams-Moulton) in
  `cppde_multistepper.hpp`, `cppde_multistepper_controller.hpp` and
  `cppde_newton.hpp` is a port of the corresponding routines in
  [SUNDIALS/CVODE(S)](https://computing.llnl.gov/projects/sundials)
  (Copyright © 2002-2024 Lawrence Livermore National Security and Southern
  Methodist University), distributed under the **BSD-3-Clause** license.
- The Rosenbrock4 stepper architecture in `cppde_rosenbrock4.hpp` (and the
  surrounding stage-matrix, dense-output and stepper-protocol infrastructure)
  is derived from
  [Boost.Numeric.Odeint](https://www.boost.org/doc/libs/release/libs/numeric/odeint/)
  by Karsten Ahnert, Mario Mulansky and Christoph Koke, distributed under the
  **Boost Software License 1.0**.

Full upstream copyright notices and license texts are reproduced in
[`inst/COPYRIGHTS`](inst/COPYRIGHTS). Both upstream licenses are permissive and
compatible with MIT; using or redistributing cppDE requires preserving the
notices in `inst/COPYRIGHTS` and the relevant header attribution blocks.

The optional system libraries linked at run time are not vendored: SUNDIALS
(BSD-3-Clause) for the `cvode()` backend and SuiteSparse-KLU (LGPL-2.1+) for
the sparse-Jacobian path. Their own licenses govern the libraries at the
user's installation site.
