---
output:
  github_document:
    toc: false
    html_preview: false
---

<!-- fitCoPt.md is generated from fitCoPt.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/fitCoPt-",
  fig.width = 9,
  fig.height = 4,
  out.width = "100%",
  dev = "svglite",
  message = FALSE,
  warning = FALSE
)
```

# Fitting measured magnetisation dynamics

Fit of a Landau-Lifshitz-Gilbert model to the pump-probe measurements of
Zhang et al. (2025). cppDE integrates the model and its parameter
sensitivities. Likelihood, noise model and optimiser are R functions from
[`inst/examples/fitTools.R`](../inst/examples/fitTools.R). The code chunks are the
sections of
[`inst/examples/example_fitCoPt.R`](../inst/examples/example_fitCoPt.R).

```{r example-sections, include = FALSE}
## Chunks below are read from the numbered sections of the example script.
example <- readLines("../inst/examples/example_fitCoPt.R")
banner  <- grep("^## -{20,}$", example)
first   <- banner[c(TRUE, FALSE)]
last    <- c(first[-1] - 1, length(example))
section <- lapply(seq_along(first), function(i) {
  code <- example[(first[i] + 3):last[i]]
  code[cumsum(nzchar(code)) > 0 & rev(cumsum(rev(nzchar(code)))) > 0]
})
names(section) <- sub("^## ([0-9]+)\\..*", "\\1", example[first + 1])
```

## The experiment

| | |
|:--|:--|
| sample | [Co (0.5 nm) / Pt (0.9 nm)]<sub>5</sub>, perpendicular magnetic anisotropy |
| pump | 80 fs, 1500 nm |
| probe | anomalous Hall conductivity, read with a THz pulse, delays up to 80 ps |
| field | 1 to 3 T, seven values, about 30° from the film normal |

The signal rises within 0.5 ps (ultrafast demagnetisation) and recovers within
a few picoseconds. The recovery carries a precession at 40 to 100 GHz. The
seven traces are fitted jointly.

```{r setup-fit}
library(cppDE)
library(ggplot2)
source(system.file("examples", "fitTools.R", package = "cppDE"))

theme_set(theme_bw(base_size = 11))
cores <- if (.Platform$OS.type == "windows") 1L else min(10L, parallel::detectCores())
```

```{r data, code = section[["1"]]}
```

## Model

The magnetisation direction $\mathbf m$, $|\mathbf m| = 1$, follows the
Landau-Lifshitz-Gilbert equation in Landau-Lifshitz form,

$$
\dot{\mathbf m} = -\frac{\gamma}{1+\alpha^2}
\left[\mathbf m\times\mathbf h + \alpha\,\mathbf m\times(\mathbf m\times\mathbf h)\right],
\qquad \gamma = \frac{g\,\mu_B}{\hbar},
$$

with Gilbert damping $\alpha$ and g-factor $g$. The effective field is the
applied field, tilted by $\phi$ from the normal $\hat{\mathbf z}$, plus the
perpendicular anisotropy field $H_a$,

$$
\mathbf h = H\,(\sin\phi,\,0,\,\cos\phi) + H_a\,\bigl(1 - k(t)\bigr)\,m_z\,\hat{\mathbf z}.
$$

Heating lowers the anisotropy field by the fraction $k(t) = K\,\kappa(t)$,
$\kappa(0) = 1$, which decays as $\dot\kappa = -\kappa/\tau_K$.

Before the pump, $\mathbf m$ rests at
$\mathbf m_0 = (\sin\theta_0, 0, \cos\theta_0)$, where the torque vanishes,

$$
H\sin(\theta_0-\phi) + H_a\sin\theta_0\cos\theta_0 = 0 ,
$$

and the field there is parallel to $\mathbf m_0$,
$\mathbf h(\mathbf m_0) = \lambda\,\mathbf m_0$ with
$\lambda = H\cos(\theta_0-\phi) + H_a\cos^2\theta_0$. The drop of $H_a$
moves the equilibrium; $\mathbf m$ precesses about it while $H_a$ recovers.

The deviation from $\mathbf m_0$ is of order $K$, and the states of the model
are $\boldsymbol\xi = (\mathbf m - \mathbf m_0)/K$ and $\kappa$. Because
$\mathbf m_0\times\mathbf h(\mathbf m_0) = 0$, the right-hand side in these
states is a polynomial in $K$ without a constant term. The equations hold for
any $K$, and $K\to 0$ gives the linearised dynamics. This avoids computing
$m_{z0} - m_z$ as a difference of nearly equal numbers.

```{r model, code = section[["2"]]}
```

## Observable, sensitivities and noise

The anomalous Hall conductivity is proportional to the out-of-plane
magnetisation. The recovery of the demagnetisation is not part of the model and
is described per measurement by two exponentials:

$$
y(t) = A_1 e^{-t/\tau_1} + A_2 e^{-t/\tau_2} + S\,\bigl[m_{z0} - m_z(t)\bigr]
     = A_1 e^{-t/\tau_1} + A_2 e^{-t/\tau_2} - S\,K\,\xi_z(t).
$$

$\alpha$, $g$, $H_a$, $\tau_K$, $S$, $K$ and $\phi$ are shared by all fields;
background and noise level are per measurement. $H$ is the set field value;
$\mu_B/\hbar$ is the only fixed constant.

cppDE returns $\partial\boldsymbol\xi/\partial p$ for the parameters of the
model. $\theta_0$ and $\lambda$ depend on $H_a$ and $\phi$; the implicit
function theorem gives

$$
\frac{\partial\theta_0}{\partial H_a} = -\frac{\sin\theta_0\cos\theta_0}{D},\qquad
\frac{\partial\theta_0}{\partial\phi} = \frac{H\cos(\theta_0-\phi)}{D},\qquad
D = H\cos(\theta_0-\phi) + H_a\cos 2\theta_0 ,
$$

and the chain rule gives the sensitivities with respect to $H_a$ and $\phi$.
No finite differences are used.

The autocorrelation of the residuals decays within five samples, about one
picosecond. The noise is modelled as a moving average (Box et al. 2015),

$$
u_t = \eta_t + \sum_{j=1}^{5} w_j\,\eta_{t-j},\qquad \eta_t\sim\mathcal N(0,\sigma^2),
$$

with the kernel $w$ shared by all measurements and $\sigma$ per measurement.
Correlations beyond five samples are zero, so the noise model cannot describe
the precession. The likelihood is the conditional sum of squares of the
innovations $\eta_t$; the $\log\sigma$ term enters as a pseudo-residual, which
keeps the problem least squares.

```{r likelihood, code = section[["3"]]}
```

## Fit

Gauss-Newton on a trust region with exact subproblem (Moré-Sorensen), from
100 random starts. Positive parameters are fitted as $\log_{10}$, the fractions
$K$ and $\phi/90^\circ$ as $\log_{10}\bigl(x/(1-x)\bigr)$, so one unit is one
decade. The physics parameters carry a Gaussian prior with a standard deviation
of three decades, centred on orders of magnitude and the nominal field angle.

```{r multistart, code = section[["4"]]}
```

```{r fit, code = section[["5"]]}
```

The innovations pass the Ljung-Box test (Ljung and Box 1978). For small precession amplitudes, $S$,
$K$ and $\phi$ enter to first order as a product and are not determined
individually.

Small-angle precession frequency about the equilibrium (Smit-Beljers) against
the frequencies obtained in the paper from damped-sine fits:

```{r frequencies, code = section[["7"]]}
```

Data: measured traces of Fig. 3 in Z. Zhang, Y. Shiota, S. Karube,
Y. Watanabe, T. Ono and H. Hirori,
[Phys. Rev. Applied **24**, 044079 (2025)](https://doi.org/10.1103/4p39-gyrd),
published under CC BY 4.0 in
[doi:10.5281/zenodo.17333041](https://doi.org/10.5281/zenodo.17333041); see
[`inst/COPYRIGHTS`](../inst/COPYRIGHTS).

Noise model: G. E. P. Box, G. M. Jenkins, G. C. Reinsel and G. M. Ljung,
*Time Series Analysis: Forecasting and Control*, 5th ed. (Wiley, 2015).
G. M. Ljung and G. E. P. Box, *On a measure of lack of fit in time series
models*, [Biometrika **65**, 297 (1978)](https://doi.org/10.1093/biomet/65.2.297).
